### 2013-07-01 Publications/2013-TMI/04-methods.tex
Updating numerical implementation section
* Implemented #80, but still missing the consecuences of p(U) in the general framework
* Re-written methodology in a fashion a bit more consistent with the undetermined number of regions

All the surveyed frameworks are deduced from the general level set evolution equation
introduced by \citep{osher_fronts_1988}:
\begin{equation}
\frac{\partial \Phi_D(x,t)}{\partial t} = \nu ( \Phi_D(x,t)) \, \left| \nabla \Phi_D(x,t) \right|
\end{equation}
where $\nu$ is the velocity of the flow or speed function that contains the local
segmentation and contour regularization constraints, and $\Phi_D: \Omega \to \mathbb{R}$
is the signed distance function often used to represent implicitly the active contour
by its zero level. An early extension to image registration was proposed by 
\citep{vemuri_joint_2003} but replacing $\Phi_D$ by the intensity function of the
image to register, $\Phi_I: \Omega \to \mathbb{R}$ (the moving image). 
\citep{bertalmio_morphing_2000,vemuri_joint_2003} then define the speed function as
$\nu(\Phi_I(x,t)) = \Phi_I(x,t) - \Phi_T(x,t)$, where $\Phi_T(x)$ is the intensity
function of the target image. Finally, \citep{gorthi_active_2011} propose a different
multiphase label function as level set function ($\Phi_L$) that improves registration
results.

Naming $\Phi_G$ a generic objective function, a dense deformation field of vectors 
$u: \mathbb{R}^n \to \mathbb{R}^n$ (tipically $n = \{ 2, 3 \}$) is introduced. 
Then, the conservation of the morphological description is assumed, such as
$\Phi_G(x,t) = \Phi_G( x + du, t + dt ) \implies d\Phi_G(x,t) = 0$, where $d\Phi_G$ is
the total derivative of $\Phi_G$. Using the chain rule, it can be rewritten as the 
evolution equation of a vector flow:
\begin{equation}
\frac{\partial u(x,t)}{\partial t}= - \frac{\Phi_G}{\left| \nabla \Phi_G \right|} N_{\Phi_G},
\end{equation}
where $N_{\Phi_G}$ is the normal of the level set. Finally, they introduce the general
level set evolution equation into the vector flow to perform the registration task:
\begin{equation}
\frac{\partial u(x,t)}{\partial t} = - \nu( \Phi_G(x,t) ) N_{\Phi_G}.
\end{equation}
Therefore, the position of the level set function $\Phi_G$ at time $t$ is given by the
deformation field $u(x,t)$ and the initial level set function $\Phi_G(x,0)$, yielding:
\begin{equation}
\frac{\partial u(x,t)}{\partial t} = - \nu( \Phi_G(x + u(x,t), 0) ) N_{\Phi_G}.
\end{equation}

@@

\subsection{From the Maximum A-Posteriori Model to the Mumford-Shah functional}
\label{sec:methods_map}
%
A widely used approach to image segmentation is derived from the
Bayes' rule \eqref{eq:bayes_rule}, where one seeks for a partitioning
of a certain image $X$ in piecewise smooth regions $\Omega = \lbrace \Omega_k , 
k\in\left[ 1 .. K \right] \rbrace$,  that maximizes the a posteriori 
probability given the multivariate image $X \in \mathbb{R}^C$, 
with $C$ being the number of image channels.
\begin{equation}
p(Y \mid X) = \frac{p(X \mid Y)\, p(Y)}{p(X)}.
\label{eq:bayes_rule}
\end{equation}

Therefore, $Y$ is a certain realization of the piecewise 
disjoint region set $\Omega$. $p(X \mid Y)$ is the \emph{likelihood} of 
the realization of $X$ (the image) given a certain distribution model for 
each region $\Omega_k$. The second term, $p(Y)$, is the a-priori probability of 
the partitioning $Y$. Finally, $p(X)$ is the probability of a certain image 
realization, and thus, it will remain constant when computing the \gls{map}.
Consequently, $p(Y \mid X) \propto p(X \mid Y)\, p(Y)$, and:
\begin{equation}
\underset{Y}{\argmax} \left\{ p(Y \mid X) \right\} = 
\underset{Y}{\argmax} \left\{ p(X \mid Y)\, p(Y) \right\}.
\end{equation}

An extended assumption is that the feature vector realization $X$ is
\emph{i.i.d.}, and thus, it is possible to write the a-posteriori
probability $p(X \mid Y)$ as a continuous product with $d\mathbf{x}$ the
infinitesimal voxel size:
\begin{equation}
p(X \mid Y) \, p(Y) = \underset{k}{\prod} \underset{\mathbf{x}\in \Omega_k}{\prod}
p_k( X(\mathbf{x}) \mid Y(\mathbf{x}) )^{d\mathbf{x}},
\label{eq:bayes_aposteriori}
\end{equation}
where the prior probability $p(Y)$ is implicitly
defined by the regions definition.

A second widely-accepted assumption is the multivariate normal 
distribution of the different tissues in \gls{mri} data. Therefore,
the posterior probability of an infinitesimal voxel
can be written as:
\begin{equation}
p_k( X(\mathbf{x}) \mid Y(\mathbf{x}) ) = \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{k}\right|}}\,{e^{\left(-\frac{1}{2}  \Delta^2_k (\mathbf{X(\mathbf{x})}) \right)}}.
\label{eq:bayes_mpdf}
\end{equation}
where we can identify the factor in the exponential, with 
$\mathbf{f} = X(\mathbf{x})$ as the squared \emph{Mahalanobis 
distance} with the parameters set $\Theta_k = \lbrace \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k \rbrace$:
\begin{equation}
\Delta^2_k (\mathbf{f} \mid \Theta_k ) = (\mathbf{f} - \boldsymbol{\mu}_k)^T \, \boldsymbol{\Sigma}^{-1}_k \, (\mathbf{f} - \boldsymbol{\mu}_k).
\label{eq:bayes_mahalanobis}
\end{equation}

Finally, we can turn the \gls{map} problem into a variational one
applying the following log-transform:
\begin{multline}
E(X \mid Y)= -\log \left[ p(X \mid Y) \, p(Y) \right] = \\
= -\log \left[ \underset{k}{\prod} \underset{\mathbf{x}\in \Omega_k}{\prod}
p_k( X(\mathbf{x}) \mid Y(\mathbf{x}) )^{d\mathbf{x}} \right] = \\
= \sum\limits_k \int_{\Omega_k} -\log \left[ p_k(X(\mathbf{x}) \mid Y(\mathbf{x} ) ) \right] \, d\mathbf{x},
\label{eq:energy_1}
\end{multline}
and introducing the posterior probability term \eqref{eq:bayes_mpdf}, 
we can express the functional in terms of $\lbrace\Theta_k,\Omega_k\rbrace$:
\begin{multline}
E(\Theta_k,\Omega_k) = \\
= \sum\limits_k \int_{\Omega_k} -\log \left[ \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{k}\right|}}\,{e^{\left(-\frac{1}{2}  \Delta^2_k (\mathbf{f}) \right)}} \right] \, d\mathbf{x} = \\
= \sum\limits_k \int_{\Omega_k} \left[ \frac{1}{2} \log{ \left( (2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{k}\right| \right)} + \frac{1}{2}  \Delta^2_k (\mathbf{f}) \right] \,d\mathbf{x}.
\end{multline}
Finally, after removing scaling factors and independent constants,
we obtain:
\begin{align}
E(\Theta_k,\Omega_k) = \sum\limits_k \int_{\Omega_k} \left[ \log \left|\mathbf{\Sigma}_k\right| + \Delta^2_k (\mathbf{f}) \right] \,d\mathbf{x},
\label{eq:map_energy}
\end{align}
\todo[inline]{improve this explanation} where we have a constant term scaled by
the total volume of the partition $\Omega_k$ and the determinant of the 
covariance matrix of the partition $\left|\boldsymbol{\Sigma}_{k}\right|$,
plus an energy term based on the squared \emph{Mahalanobis distance}.

Equation \eqref{eq:map_energy} resembles the Mumford-Shah functional including variance,
that modifies the original functional in a way that it can deal with more general
distributions. This is necessary to avoid the assumption that regions $\Omega_k$
have a fixed covariance matrix on their complete domain. One immediate advantage
of this functional from the original one is the possibility to distinguish regions
with the same mean vector but different covariance matrix \citep{brox_local_2009}:
\begin{multline}
E(\Theta_k,\Omega_k) = \sum\limits_k \int_{\Omega_k} \left[ \log \left|\mathbf{\Sigma}_k\right| + \Delta^2_k (\mathbf{f}) \right] \,d\mathbf{x} \\
+ \lambda \int_{\Omega_k - K}  ( \left| \nabla \mathbf{\mu} \right| ^2 + \left| \nabla \mathbf{\Sigma}_k \right| ^2 ) \, d\mathbf{x} 
+ \nu |K|,
\end{multline}
that is easily identifiable with \eqref{eq:map_energy} when we apply 
the so-called \emph{cartoon limit}, 
for $\lambda \to \infty$:
\begin{equation}
E(\Theta_k,K) = \sum\limits_k \int_{\Omega_k} \left[ \log \left|\mathbf{\Sigma}_k\right| + \Delta^2_k (\mathbf{f}) \right] \,d\mathbf{x}
+ \nu |K|.
\end{equation}

@@

\subsection{Deformation model}
\label{sec:deformation}
%
The segmentation problem is transformed into a registration one if the initial
partition $Y$ is derived from the initial shape-priors in reference space. 
Introducing a dense deformation field $u$ that maps the original partition to
a better fit of the regions in the target coordinate space. Thus, the minimization
problem stated in \eqref{eq:map_energy} can be expressed so that $u$ is the 
unknown:
\begin{equation}
E(u(\mathbf{x})) = \sum\limits_k \int_{\Omega'_k} \left[ \log \left|\mathbf{\Sigma}_k\right| + \Delta^2_k (\mathbf{f'}) \right] \,d\mathbf{x}
\label{eq:map_energy_deformation}
\end{equation}
where $\Omega'_k = u(\Omega_k)$ and $\mathbf{f'} = X(u(\mathbf{x}))$.

\subsection{\Acrlong{acwe} based segmentation model}
%
Let us denote $\{c_i\}_{i=1..N_c}$ the nodes of a shape prior surface. In
our application, a precise \gls{wm}-\gls{gm} interface extracted from a
high-resolution reference volume. All the formulations can be naturally
extended to include more shape priors. On the other hand, we have a 
number of \gls{dwi}-derived features at each
voxel of the volume. Let us denote by $x$ the voxel and 
$f(x) = [ f_1, f_2, \ldots, f_N]^T(x)$ its associated feature vector.\\
%
The transformation from reference into \gls{dwi} coordinate space is 
achieved through a dense deformation field $u(x)$, such that:
%
\begin{equation}
c_i' = T\{c_i\} = c_i + u(c_i)
\end{equation}
% 
Since the nodes of the anatomical surfaces might lay off-grid, it is 
required to derive $u(x)$ from a discrete set of parameters $\{u_k\}_{k=1..K}$.
Densification is achieved through a set of associated basis functions 
$\psi_k$ (e.g. rbf, interpolation splines):
%
\begin{equation}
u(x) = \sum_k \psi_k(x) u_k
\end{equation}
%
Consequently, the transformation writes
%
\begin{equation}
\label{eq:transformation}
c_i' = T\{c_i\} = c_i + u(c_i) = c_i + \sum_k \psi_k(c_i)u_k
\end{equation} 
%
% Comment: maybe this is not for IPMI 2013.
%Note that, since $c_i$ remains constant in the DW segmentation process,
%the values of $\psi_k(c_i)$ can be precomputed. Also, provided compact 
%support of the basis functions, the system remains relatively sparse.\\
%
Based on the current estimate of the distortion $u$, we can compute 
``expected samples'' within the shape prior projected into the \gls{dwi}.
Thus, we now estimate region descriptors of the \gls{dwi} features 
$f(x)$ of the regions defined by the priors in \gls{dwi} space.
%
Using Gaussian distributions as region descriptors, we propose an
\gls{acwe}-like, piece-wise constant, variational image segmentation
model (where the unknown is the deformation field)
\cite{chan_active_2001}:

where $R$ indexes the existing regions and the integral domains
depend on the deformation field $u$. Note
that minimizing this energy, $\argmin_u\{E\}$, yields the \gls{map} 
estimate of a piece-wise smooth image model affected by Gaussian 
additive noise. This inverse problem is ill-posed
\cite{bertero_ill-posed_1988,hadamard_sur_1902}.
In order to account for deformation field regularity and to render the 
problem well-posed, we include limiting and regularization terms into 
the energy functional \cite{morozov_linear_1975,tichonov_solution_1963}:
%
\begin{align}
\label{eq:complete_energy}
E(u) &= \sum_{\forall{R}} \lbrace \int_{\Omega_R} (f-\mu_R)^T\Sigma_R^{-1}(f-\mu_R) dx \rbrace \nonumber \\
&\quad + \alpha \int  \|u\|^2 dx + \beta \int \left( \|\nabla u_x\|^2 + \|\nabla u_y\|^2 + \|\nabla u_z\|^2\right) dx
\end{align}
%
These regularity terms ensure that the segmenting contours in 
\gls{dwi} space are still close to their native shape. The model
easily allows to incorporate inhomogeneous and anisotropic 
regularization \cite{nagel_investigation_1986} to better regularize
the \gls{epi} distortion. \\
%

At each iteration, we update the distortion along the steepest 
energy descent. This gradient descent step can be efficiently 
tackled by discretizing the time in a forward Euler scheme, 
and making the right hand side semi-implicit in the 
regularization terms:
%
\begin{align}
\frac{u^{t+1}-u^t}{\tau} &= - \sum_{i=1}^{N_c} \left[ e(f(c_i'))  \hat{n}_{c_i'} \psi_{c_i}(x) \right] -\alpha u^{t+1} + \beta\Delta u^{t+1}
\end{align}
%
where the data terms remain functions of the current estimate 
$u^t$, thus $c_i' = c_i'(u^t)$. For simplicity on notation, we 
restricted the number of priors to only 1. We also defined 
$e(f(c_i')) = E_{out}(f(c_i')) - E_{in}(f(c_i'))$, 
and $E_R(f) = {(f-\mu_R)^T\Sigma_R^{-1}(f-\mu_R)}$.
We applied a spectral approach to solve this implicit scheme:
%
\begin{equation}
u^{t+1} = \mathcal{F}^{-1}\left\{ \frac{\mathcal{F}\{u^t/\tau
- \sum_{i=1}^{N_c} \left[e(f(c_i')) \hat{n}_{c_i'} \psi_{c_i}(x) \right]  \}}{\mathcal{F}\{(1/\tau+\alpha)\mathcal{I}-\beta\Delta\}} \right\}
\end{equation}
%

\subsection{Operator splitting}
In order to make the Energy minimisation computationally more tractable, we propose the following operator splitting: Let us optimize the data terms and the regularity terms on separate copies of the deformation field, now called $u$ and $v$, constrained to be equal:
\begin{align}
E(u,v) &= \int_{w'(u)} (f-\mu_w)^T\Sigma_w^{-1}(f-\mu_w) dx\nonumber\\
&\quad +\int_{g'(u)} (f-\mu_g)^T\Sigma_g^{-1}(f-\mu_g) dx\\
&\quad +\int_{o'(u)} (f-\mu_o)^T\Sigma_o^{-1}(f-\mu_o) dx\nonumber\\
&\quad +\int v^T A v dx + \int \tr\{(\nabla v^T)^T B (\nabla v^T)\} dx\nonumber
\end{align}
and now
\begin{equation}
\min_{u,v} \{ E \} \quad s.t. \quad u = v
\end{equation}
In order to take the equality constraint into account, we may make use of augmented Lagrangians (a combination of Lagrangian multipliers and penalty terms on the constraint)
\cite{bertsekas_multiplier_1976,glowinski_augmented_1989,nocedal_numerical_2006}:
\begin{equation}
AL(u,v,\lambda,r) = E(u,v) + \langle \lambda, u-v \rangle + \frac{r}{2} \| u - v \|_2^2
\end{equation}
To solve the constraint minimization problem, we may now optimize the Augmented Lagrangian in an iterative way:
\begin{equation}
\left\lbrace \begin{array}{rcl}
u^{t+1} &=& \argmin_{u} AL(u,v^t,\lambda^t,r)\\
v^{t+1} &=& \argmin_{v} AL(u^{t+1},v,\lambda^t,r)\\
\lambda^{t+1} &=& \lambda^t + \rho(u^{t+1}-v^{t+1})
\end{array}\right.\end{equation}
where typically $0 < \rho < r$. The two subminimization problems will now be much easier to handle than the original complete problem (``divide et impera'').

\subsection{Shape gradients}
To compute the gradient-descent of the data-term domain integrals with respect to the underlying deformation field, we want to make use of shape gradients \cite{jehan-besson_dream2s:_2003,herbulot_segmentation_2006}. A little bit of theory is therefore in order.

>>>>

All the surveyed frameworks are deduced from the general level set evolution equation
introduced by \citep{osher_fronts_1988}:
\begin{equation}
\frac{\partial \Phi_D(\mathbf{x},t)}{\partial t} = \nu ( \Phi_D(\mathbf{x},t)) \, \left| \nabla \Phi_D(\mathbf{x},t) \right|
\end{equation}
where $\nu$ is the velocity of the flow or speed function that contains the local
segmentation and contour regularization constraints, and $\Phi_D: \Omega \to \mathbb{R}$
is the signed distance function often used to represent implicitly the active contour
by its zero level. An early extension to image registration was proposed by 
\citep{vemuri_joint_2003} but replacing $\Phi_D$ by the intensity function of the
image to register, $\Phi_I: \Omega \to \mathbb{R}$ (the moving image). 
\citep{bertalmio_morphing_2000,vemuri_joint_2003} then define the speed function as
$\nu(\Phi_I(\mathbf{x},t)) = \Phi_I(\mathbf{x},t) - \Phi_T(\mathbf{x},t)$, where $\Phi_T(\mathbf{x})$ is the intensity
function of the target image. Finally, \citep{gorthi_active_2011} propose a different
multiphase label function as level set function ($\Phi_L$) that improves registration
results.

Naming $\Phi_G$ a generic objective function, a dense deformation field of vectors 
$u: \mathbb{R}^n \to \mathbb{R}^n$ (tipically $n = \{ 2, 3 \}$) is introduced. 
Then, the conservation of the morphological description is assumed, such as
$\Phi_G(\mathbf{x},t) = \Phi_G( \mathbf{x} + du, t + dt ) \implies d\Phi_G(\mathbf{x},t) = 0$, where $d\Phi_G$ is
the total derivative of $\Phi_G$. Using the chain rule, it can be rewritten as the 
evolution equation of a vector flow:
\begin{equation}
\frac{\partial u(\mathbf{x},t)}{\partial t}= - \frac{\Phi_G}{\left| \nabla \Phi_G \right|} N_{\Phi_G},
\end{equation}
where $N_{\Phi_G}$ is the normal of the level set. Finally, they introduce the general
level set evolution equation into the vector flow to perform the registration task:
\begin{equation}
\frac{\partial u(\mathbf{x},t)}{\partial t} = - \nu( \Phi_G(\mathbf{x},t) ) N_{\Phi_G}.
\end{equation}
Therefore, the position of the level set function $\Phi_G$ at time $t$ is given by the
deformation field $u(\mathbf{x},t)$ and the initial level set function $\Phi_G(\mathbf{x},0)$, yielding:
\begin{equation}
\frac{\partial u(\mathbf{x},t)}{\partial t} = - \nu( \Phi_G(\mathbf{x} + u(\mathbf{x},t), 0) ) N_{\Phi_G}.
\end{equation}

@@

\subsection{Bridging segmentation frameworks}
\label{sec:methods_map}
%
A widely used approach to image segmentation is derived from the
Bayes' rule \eqref{eq:bayes_rule}. This framework seeks for a 
partitioning of a certain observation of features $F: \mathbb{R}^n \to %
\mathbb{R}^C$ (an $n$-dimensional \emph{image} comprising $C$ different 
channels) in piecewise smooth regions $\Omega = \lbrace \Omega_k , 
k\in\left[ 1 \ldots K \right] \rbrace$,  that maximize the \emph{a posteriori}
probability $p(\Omega \mid F)$:
\begin{equation}
p(\Omega \mid F) = \frac{p(F \mid \Omega)\, p(\Omega)}{p(F)},
\label{eq:bayes_rule}
\end{equation}
where $p(F \mid \Omega)$ is the \emph{likelihood} of the realization 
of $F$ given a certain region set $\Omega$. The second term, $p(\Omega)$,
is the a-priori probability of the partitioning. Finally, $p(F)$ is the 
probability of a certain image realization, and thus, it will remain 
constant when computing the \gls{map}. Consequently, the Bayes' rule
can be simplified for the optimization as follows:
\begin{equation}
p(\Omega \mid F) \propto p(F \mid \Omega)\, p(\Omega).
\label{eq:bayes_rule_simplified}
\end{equation}

In the simplest Bayesian segmentation framework, the unknown is the
parameter set $\Theta_k$ of region-wise descriptors that minimize 
\eqref{eq:bayes_rule}. Conversely, we can assume these descriptors as
fixed (fulfilling the concept of conservation of morphological description
mentioned before) and introduce a realization of a dense deformation field $U: %
\mathbb{R}^n \to \mathbb{R}^n$ that maps the initial partition $\Omega_0$
to the estimated one:
\begin{equation}
\Omega' = \Omega_0 + \hat{U}.
\label{eq:omega_tf}
\end{equation}

Thus, \eqref{eq:bayes_rule_simplified} is rewritten as 
$p(U \mid F) \propto p(F \mid U)\, p(U)$ that yields the
following \gls{map} solution:

\begin{equation}
\hat{U} = \underset{U}{\argmax} \left\{ p(U \mid F) \right\} = 
\underset{U}{\argmax} \left\{ p(F \mid U)\, p(U) \right\}.
\label{eq:map_u}
\end{equation}

An extended assumption is that the feature vector realization $F$ is
\emph{i.i.d.}, and thus, it is possible to write the a-posteriori
probability $p(F \mid U)$ as a continuous product with $d\mathbf{x}$ the
infinitesimal voxel size:
\begin{equation}
p(F \mid U) \, p(U) = \underset{k}{\prod} \underset{\mathbf{x}\in \Omega'_k}{\prod}
p_k\left( F(\mathbf{x} + U(\mathbf{x})) \mid U(\mathbf{x}) \right)^{d\mathbf{x}}.
\label{eq:bayes_aposteriori}
\end{equation}
\todo[inline]{I'm not convinced about turning $p(\Omega)$ into $p(U)$.}

A second widely-accepted assumption is the multivariate normal 
distribution of the different tissues in \gls{mri} data. Therefore,
the posterior probability of an infinitesimal voxel can be written as:
\begin{equation}
p_k( F(\mathbf{x}) \mid U(\mathbf{x}) ) = \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{k}\right|}}\,{e^{\left(-\frac{1}{2}  \Delta^2_k (\mathbf{f}') \right)}}.
\label{eq:bayes_mpdf}
\end{equation}
where we can identify the factor in the exponential as the squared \emph{Mahalanobis 
distance} of the feature observed in the displaced position, $\mathbf{f}' = 
F(\mathbf{x} + U(\mathbf{x}))$ with region descriptors
$\Theta_k = \lbrace \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k \rbrace$:
\begin{equation}
\Delta^2_k (\mathbf{f} \mid \Theta_k ) = (\mathbf{f} - \boldsymbol{\mu}_k)^T \, \boldsymbol{\Sigma}^{-1}_k \, (\mathbf{f} - \boldsymbol{\mu}_k).
\label{eq:bayes_mahalanobis}
\end{equation}

Finally, we can turn the \gls{map} problem into a variational one
applying the following log-transform:
\begin{multline}
E(F \mid U)= -\log \left[ p(F \mid U) \, p(U) \right] = \\
= -\log \left[ \underset{k}{\prod} \underset{\mathbf{x}\in \Omega'_k}{\prod}
p_k( F(\mathbf{x}) \mid U(\mathbf{x}) )^{d\mathbf{x}} \right] = \\
= \sum\limits_k \int_{\Omega'_k} -\log \left[ p_k(F(\mathbf{x}) \mid U(\mathbf{x} ) ) \right] \, d\mathbf{x},
\label{eq:energy_1}
\end{multline}
and introducing the posterior probability term \eqref{eq:bayes_mpdf}, 
we can express the functional in terms of the deformation field $U$ and
the region descriptors $\Theta_k$:
\begin{multline}
E(U) = \\
= \sum\limits_k \int_{\Omega'_k} -\log \left[ \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{k}\right|}}\,{e^{\left(-\frac{1}{2}  \Delta^2_k (\mathbf{f}) \right)}} \right] \, d\mathbf{x} = \\
= \sum\limits_k \int_{\Omega'_k} \left[ \frac{1}{2} \log{ \left( (2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{k}\right| \right)} + \frac{1}{2}  \Delta^2_k (\mathbf{f}) \right] \,d\mathbf{x}.
\end{multline}
Finally, after removing scaling factors and independent constants,
we obtain:
\begin{align}
E(U) = \sum\limits_k \left[ \left|\Omega'_k\right|\,\log \left|\mathbf{\Sigma}_k \right| + \int_{\Omega'_k} \Delta^2_k (\mathbf{f}) \,d\mathbf{x} \right],
\label{eq:map_energy}
\end{align}
where we have a constant term scaled by the total volume of the partition $\Omega_k$ and 
the determinant of the  covariance matrix of the partition $\left|\boldsymbol{\Sigma}_{k}\right|$,
plus an energy term based on the squared \emph{Mahalanobis distance}.
As long as this shape-based registration framework is designed to allow small changes
between boundaries (and therefore, their volumes), and the shape descriptors should not
change significantly due to the conservation principle, the first term can be optionally
dismissed with respect the second one, what yields:
\begin{align}
E(U) = \sum\limits_k \int_{\Omega'_k} \Delta^2_k (\mathbf{f}) \,d\mathbf{x}.
\label{eq:final_map_energy}
\end{align}

This latter expression resembles the Mumford-Shah functional 
\citep{mumford_optimal_1989} which is at the background of all \gls{acwe}-based
methods. In this case, \eqref{eq:final_map_energy} includes covariance as
region descriptor, what modifies the original functional in a way that it 
can deal with more general distributions. This is necessary to avoid the 
assumption that regions $\Omega_k$ have a fixed covariance matrix on their 
complete domain. One immediate advantage of this functional from the original 
one is the possibility to distinguish regions with the same mean vector but 
different covariance matrix \citep{brox_local_2009}. In this work, the 
Mumford-Shah functional is derived for this extension as follows:
\begin{multline}
E(\Theta_k,\Omega_k) = \sum\limits_k \int_{\Omega_k} \left[ \log \left|\mathbf{\Sigma}_k\right| + \Delta^2_k (\mathbf{f}) \right] \,d\mathbf{x} \\
+ \lambda \int_{\Omega_k - K}  ( \left| \nabla \mathbf{\mu} \right| ^2 + \left| \nabla \mathbf{\Sigma}_k \right| ^2 ) \, d\mathbf{x} 
+ \nu |K|,
\end{multline}
that is easily identifiable with \eqref{eq:map_energy} when we apply 
the so-called \emph{cartoon limit}, 
for $\lambda \to \infty$:
\begin{equation}
E(\Theta_k,K) = \sum\limits_k \int_{\Omega_k} \left[ \log \left|\mathbf{\Sigma}_k\right| + \Delta^2_k (\mathbf{f}) \right] \,d\mathbf{x}
+ \nu |K|.
\end{equation}

@@

\subsection{Introducing priors}
\label{sec:priors}
%
By introducing $U$ in \eqref{eq:bayes_rule_simplified}, we transformed 
the segmentation problem into a registration one provided with $\Omega_0$ 
(the initial partition) is derived from the shape-priors given in reference 
space. Therefore, the priors can be expressed in terms of the deformation
field:
\begin{equation}
P(U) = \underset{\mathbf{x}}{\prod}\,p(U(\mathbf{x}))\,^{d\mathbf{x}}.
\label{eq:prior_u}
\end{equation}
We define $P(U(\mathbf{x})) = p(\mathbf{u}) = \underset{d}\prod \, p_d(\mathbf{u})$ 
with $d$ the order of the derivative. We also assume that an initial
affine registration let us consider that the deformation field ($d=0$)
and its derivatives up to some order ($d>0$) are all zero mean, and
have some variance $\Sigma_d$. We restrict ourselves to $d=\lbrace 0,1 \rbrace$
(the field and its gradient):
\begin{subequations}
\begin{equation}
p_0(\mathbf{u}) = \mathcal{N}( \mathbf{u} \mid 0, \alpha^{-1}),
\end{equation}
\begin{equation}
p_1(\mathbf{u}) = \mathcal{N}( \nabla \mathbf{u} \mid 0, {\Sigma_\beta}^{-1}), 
\end{equation}
\end{subequations}
And therefore:
\begin{equation}
P(U) = \underset{\mathbf{x}}\prod \left[ p_0(\mathbf{u}) \, p_1(\mathbf{u}) \right]^{d\mathbf{x}}.
\end{equation}

\subsection{Numerical Implementation}
%
Let us denote $\{c_i\}_{i=1 \ldots N_c}$ the nodes of one or several shape-prior
surface(s). In our application, precise tissue interfaces of interest 
extracted from a high-resolution, anatomically correct reference volume. 
On the other hand, we have a number of \gls{dwi}-derived features at each
voxel of the volume. Let us denote by $\mathbf{x}$ the voxel and $f(\mathbf{x}) = 
[ f_1, f_2, \ldots, f_N]^T(\mathbf{x})$ its associated feature vector.

\subsubsection{Deformation field}
\label{sec:deformation_field}
The transformation from reference into \gls{dwi} coordinate space is 
achieved through a dense deformation field $u(\mathbf{x})$, such that:
\begin{equation}
c_i' = T\{c_i\} = c_i + u(c_i),
\end{equation}
what is equivalent to \eqref{eq:omega_tf}. Since the nodes of the anatomical 
surfaces might lay off-grid, it is required to derive $u(\mathbf{x})$ from a discrete 
set of parameters $\{u_k\}_{k=1 \ldots K}$. Densification is achieved through 
a set of associated basis functions $\psi_k$ (e.g. rbf, interpolation splines):
%
\begin{equation}
u(\mathbf{x}) = \sum_k \psi_k(\mathbf{x}) u_k
\end{equation}
%
In our implementation, $\psi_k$ is chosen to be a tensor-product B-Spline kernel.
Then, the transformation writes
%
\begin{equation}
\label{eq:transformation}
c_i' = T\{c_i\} = c_i + u(c_i) = c_i + \sum_k \psi_k(c_i)u_k
\end{equation}

Based on the current estimate of the distortion $u$, we can compute 
``expected samples'' within the shape prior projected into the \gls{dwi}.
Thus, we now estimate region descriptors of the \gls{dwi} features 
$f(\mathbf{x})$ of the regions defined by the priors in \gls{dwi} space.
%
Using the region descriptors derived in \autoref{sec:methods_map}, we propose
an \gls{acwe}-like, piece-wise constant, variational image segmentation
model (where the unknown is the deformation field)
\cite{chan_active_2001} with the energy functional obtained in 
\eqref{eq:final_map_energy}. This inverse problem is ill-posed
\cite{bertero_ill-posed_1988,hadamard_sur_1902}.
In order to account for deformation field regularity and to render the 
problem well-posed, we include limiting and regularization terms into 
the energy functional \cite{morozov_linear_1975,tichonov_solution_1963}:
%
\begin{multline}
E(u) = \sum\limits_k \int_{\Omega'_k} \Delta^2_k (\mathbf{f}) \,d\mathbf{x} \\
+ \int u^T \, A \, u \, d\mathbf{x} + \int \tr\{(\nabla u^T)^T B (\nabla u^T)\} d\mathbf{x}
\label{eq:complete_energy}
\end{multline}
%
These regularity terms ensure that the segmenting contours in 
\gls{dwi} space are still close to their native shape. The model
easily allows to incorporate inhomogeneous and anisotropic 
regularization \cite{nagel_investigation_1986} to better regularize
the \gls{epi} distortion. \\
%

\subsubsection{Operator splitting}
In order to make the energy minimization computationally more tractable, 
we propose the following operator splitting: Let us optimize the data terms 
and the regularity terms on separate copies of the deformation field, 
now called $u$ and $v$, constrained to be equal:
\begin{multline}
E(u,v) = \sum\limits_k \int_{\Omega'_k(u)} \Delta^2_k (\mathbf{f}) \,d\mathbf{x} \\
+ \int v^T \, A \, v \, d\mathbf{x} + \int \tr\{(\nabla v^T)^T B (\nabla v^T)\} d\mathbf{x}
\label{eq:operator_splitting}
\end{multline}
and now
\begin{equation}
\min_{u,v} \{ E \} \quad s.t. \quad u = v.
\end{equation}

In order to take the equality constraint into account, we may make use of 
augmented Lagrangians (a combination of Lagrangian multipliers and penalty 
terms on the constraint) \cite{bertsekas_multiplier_1976,
glowinski_augmented_1989,nocedal_numerical_2006}:
\begin{equation}
AL(u,v,\lambda,r) = E(u,v) + \langle \lambda, u-v \rangle + \frac{r}{2} \| u - v \|_2^2
\end{equation}

To solve the constraint minimization problem, we may now optimize the 
Augmented Lagrangian in an iterative way:
\begin{equation}
\left\lbrace 
\begin{array}{rcl}
u^{t+1} &=& \argmin_{u} AL(u,v^t,\lambda^t,r)\\
v^{t+1} &=& \argmin_{v} AL(u^{t+1},v,\lambda^t,r)\\
\lambda^{t+1} &=& \lambda^t + \rho(u^{t+1}-v^{t+1})
\end{array}\right.
\end{equation}
where typically $0 < \rho < r$. The two sub-minimization problems will 
now be much easier to handle than the original complete problem.

\subsubsection{Shape gradients}
To compute the gradient-descent of the data-term domain integrals with 
respect to the underlying deformation field, we use shape gradients 
\cite{jehan-besson_dream2s:_2003,herbulot_segmentation_2006}. A little 
bit of theory is therefore in order.

Let $\Omega$ be an image domain and $d\Omega$ its boundary. Further, $r(\mathbf{x})$ 
is an ``arbitrary'' function over the image domain. We now derive the domain 
integral w.r.t. the contour evolution parameter $t$ (time):
\begin{equation}
\frac{\partial}{\partial t} \int_\Omega r(\mathbf{x}) d\mathbf{x} = \int_\Omega \frac{\partial r}{\partial t}(\mathbf{x}) d\mathbf{x} - \int_{d\Omega} r(\mathbf{x}) \left\langle \frac{\partial{d\Omega}}{\partial t}, N_{d\Omega}\right\rangle d\mathbf{x}
\label{eq:shape_gradients}
\end{equation}
where $\left\langle\frac{\partial{d\Omega}}{\partial t}, N_{d\Omega}\right\rangle$ is 
the projection of the boundary movement on the unit inward normal. We recall
here that equation \eqref{eq:shape_gradients} stands the bridge between the 
explicit level set formulations surveyed in \autoref{sec:methods_background} 
with the presented approach in setting up the velocity function.
